Modelling the co-evolution of trade and culture in the Roman market

### 2015-04-07 article.tex
merge of sections

\title{Model to study co-evolution of trade an culture}
\maketitle

@@

%XRC trade networks are increasingly being recognised as a key process in ancient Mediterranean
Trades are being increasingly recognised as a key process in the structure of ancient Mediterranean societies. For example, the use of table ware in the Roman empire  has been linked to the trade network~\shortcite{brughmans_connecting_2010}. Interestingly this trade network has been shown to be a market economy~\shortcite{temin_market_2001}. The work presented in this paper proposes a framework to further study the relation between trade and culture within the roman empire. This work takes place within the EPNET project~\shortcite{remesal_epnet_2014}.

%XRC this process played a role in the trajectories of cultural change of the different regions, which were interconnected by trade
Trade roads between provinces generates cultural changes due to two main effects. First, when a new connection is made between two provinces, new goods can arrive in their respective markets. The arrival of new goods naturally changes the rate of adoption of all the goods of the market CITE. Moreover, a good recently introduced in a province can be recombined with previous goods in order to create yet a new type of good, which further modify the structure of the market of the province. The second element affecting the change of culture within a province is due to the persons travelling on the roads of the trade network, each carrying its own culture. Large movements of populations may lead to changes in the cultures. For example, when Roman soldiers were stationing in a province, Roman products such as olive oil were delivered to that province CITE ROMAN DELIVERING OIL, which may have lead to the adoption of olive oil in regions that were originally not using it.

%XRC these two elements (cultural transmission and trade networks) are traditionally been studied as isolated systems.
Usually, the studies on cultural change mechanisms focus on the adoption of cultural goods such as pottery by a population. However, these studies do not take into account economical factors, such as the cost of production or acquisition of the goods. Moreover, the effect that trade may have on the frequency of goods in the population or in neighbourhood populations is not analysed. A dedicated research field studies the effect of trade mechanisms on the change of prices and the modification of goods repartition in the population. This field relies on a specific set of framework which do not take into account the results obtained on the study of cultural change mechanisms. Certainly, this division in the analysis of cultural change and trade and cultural transmission allows for a precise formalization of their inner mechanisms and therefore a good understanding of their respective dynamics.
%XRC this is wrong because blablabla
However, as it has been shown before, trades influence cultural change by modifying both the goods present in a region and the people consuming these goods.  Therefore, in order to analyse the adoption of goods within a province one must take into account both trading strategy and cultural transmission. For example, the modification of good's consumption after the rupture of a connection in the trade system is better explained by knowing the local adoption of this good (cultural change) and the prices of this good (trade).

%XRC we propose a model of coevolution to explore the deep links between trade and cultural transmission in the ancient Mediterranean
Ultimately, we aim to create a model to study the relation between cultural change and trade networks within the Roman Empire. Examples of questions that could be asked to our model are: What is the relation between distances in the trade network and cultural changes ? Did anti-conformism promote the adaptation of new products coming from remote provinces ? In order to answer these questions the model developed has to take into account both cultural effects and trade influence. More precisely, the variations in these two aspects should be modelled and should influence each other.  Therefore, in order to answer the questions targeted we need a co-evolutionary model able to take into account both trade and cultural changes.

To the best of our knowledge two works have previously proposed frameworks that could be used to model the co-evolution of trade and culture~\shortcite{bentley_specialisation_2005,macmillan_agent-based_2008}. The first proposes a relatively simple model which is limited by 2 products, a fixed behaviour of the agents and a limited variation of cultural mechanisms. The second proposes a more complete model including a large number of goods (certain being essential and the others non-essential), but suffer from a greater complexity and the absence of the notion of trade networks. Moreover, these works do not compare the dynamics obtained to the previous studies on cultural change and trading mechanisms.

Therefore, in this work, we aim for the development of a framework to study the co-evolution of culture and trades which relies on a low number of parameters. The idea behind the development of such framework is to ease the comparison of different cultural change and trade mechanisms, which means that the framework should be flexible enough to allow for the implementation of both type of mechanisms and with enough structural elements to allow the comparison of various implementations. This article will focus on the presentation of the framework, its justification and more importantly its validation by reproducing previous results obtained on cultural change, and the ability to produce trade dynamics. The results obtained when studying trade and cultural change are compared with each other.

%Organisation of the paper
\todo{write once we are more sure of what we put}
The paper is organised as follow

\section{PREVIOUS MODELLINGS OF TRADE AND CULTURE}

This section will provide a brief review of the previous approaches to model trade and culture. While non-exhaustive, this review presents the mains trends of the previous work done.

\subsection{Neutral models of cultural change}

The first aspect of our model is focused on the dynamics of cultural change. They comprise the collection of processes that promotes or inhibits the spread of information by social interaction within a population~\shortcite[3]{boyd_origin_2005}. An increasing number of social scientists are using an evolutionary framework to model these dynamics, thus fostering the development of transdisciplinary efforts designed to understand cultural evolution~\shortcite{henrich_evolution_2003}. Archaeology has a rich tradition of evolutionary studies~\shortcite{lycett_cultural_2015}, exploring concepts such as gene-culture coevolution~\shortcite{burger_absence_2007}, diffusion processes~\shortcite{fort_synthesis_2012}, phylogenetic analysis of material culture~\shortcite{obrien_cladistics_2001} or detection of rates of change in the archaeological record~\shortcite{premo_cultural_2014}.

A large percentage of these studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependence (e.g. conformism and anti-conformism). 
These analysis are often based on the theory of neutral evolution~\shortcite{neiman_stylistic_1995}, which explains variation in the distribution of traits as a function of their frequency distribution in the population. Thus, a trait is assumed to lack any selective bias and its transmission will depend on its popularity amongst the rest of individuals. The comparison of a given frequency distribution found in data against this unbiased transmission model is traditionally used as a null-hypothesis testing for selection. There exist a large number of works using this approach as the starting point to analyse patterns of cultural change in past societies~\shortcite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010,kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

Unbiased transmission generates a clear pattern of power law distributions that can be replicated with a simple random copying transmission mechanism~\shortcite{bentley_random_2004}. Within this framework, an individual will copy the traits of a randomly chosen individual with a given probability. This copy can potentially introduce some errors in the acquired trait, which account for innovation processes. The individual will in turn continue to spread these cultural traits which will be further adopted by other individuals. This basic model can be enriched by several additional processes both in the innovation~\shortcite{schillinger_copying_2014,sole_evolutionary_2013,ziman_technological_2003} and the transmissio~\shortcite{heyes_social_1994,henrich_evolution_2003}.

In particular, trade within a market system can potentially introduce a complex dynamic bias to the neutral model. Trasmission will be linked to the price of the different products. This would also affect the frequency of the product within the population, which in turn would modify its price following a co-evolutionary dynamic.

\subsection{Models including Trade}

%Hopkins 1980: structure of tax system. two hypothesis. 1)taxation increased volume of trade in the empire? 2) tax-exporting province (recently integrated province) had to export to pay tax. roman supported by many small taxes. Not sure to cite it

Previous historical and archaeological studies to model trade in the ancient Mediterranean societies have focus on descriptive approaches~\shortcite{hopkins_taxes_1980,temin_market_2001,terpstra_trade_2011,temin_economy_2006,wilson_approaches_2009,scheidel_model_2007,kessler_organization_2007}. 
These works have conclude that the Roman economy was a type of market economy where a variety of small tightly connected markets had loose connections between them~\shortcite{temin_market_2001,temin_economy_2006}. 
Based on these conclusions, further works have attempt to define more exactly the characteristics of this market such, e.g. level of intervention of the state and its impact, or the relation-ship between local and long-distance trade~\shortcite{wilson_approaches_2009}.

Few works have attempt to model the interactions between economy and cultural change in the Roman empire~\shortcite{terpstra_trade_2011,kessler_organization_2007,scheidel_model_2007}. An analysis based on historical documents has revealed the importance of economical and social institutions for the efficiency of trades of grains in the early Roman Empire~\shortcite{kessler_organization_2007}. The same type of analysis has also revealed the importance of a number of key factors (such as commercial development, population growth and geographical mobility) for the growth of income and population in the Roman Italy~\shortcite{scheidel_model_2007}. Finally, a micro-economic model has determined the importance of reputation and threat of expulsion from a trade network as mechanisms to maintain the overall efficiency of the trade network~\shortcite{terpstra_trade_2011}. These works therefore reveal the strong relation between trade and cultural change in the Roman empire. However a large percentage of the datasets used for these studies are based on material objects found in archaeological contexts such as pottery. Most of these objects are usually represented following a neutral model, but the weight of this model within a market system is not studied.

An additional limitation of the previous works is the impossibility to test the conclusions reached by the authors. The models proposed reflects only the view of the authors on the historical and archaeological data analysed. A new type of model would be necessary to go beyond the interpretation, and allow the verification of the conclusions reached. These models should be able to generate data relative to trade which are then qualitatively compared to results of analysis of archaeological data such as amphorae or other container used to carry goods. By this mean, if a factor is really important in the economical network, its removal from the model will lead to data qualitatively different from the historical data.

The Agent Based Models (ABM) methodology offers a way to produce this kind of model within a computer~\shortcite{lake_trends_2014}. An ABM is composed of a group of agents performing actions within a virtual environment. The environment and the behaviour of the agents, despite being an simpler than the real world, are designed to be analogous to their real-world counterpart. The experimenter hand designs the agents, the environment and their interaction, and observes the outcome of all interactions through time in order to compare it to the archaelogical data obtained from the real world. The goal of the experimenter is therefore to find a plausible set of assumptions used during the creation of the ABM to produce results close to what is observed in reality.

A first example of its potential is given through a model of the trade network of tableware in the roman east~\shortcite{brughmans_roman_2014}. Among the parameters tested, preliminary results of this model, show that the number of links between production site has a strong influence on the results obtained. A second more complete example proposes a detailed model of the prehispanic Pueblo societies~\shortcite{kohler_modelling_2012}. This project proposes an ABM integrating numerous effects such as food consumption, production and landscape. The questions studied are  numerous and some works are performed on the economical system and its ties with social transmissions~\shortcite{kobti_emergence_2006,cockburn_simulating_2013}. However the work done in this project consider a pre-market society in a specific village which makes it difficult to reach more general conclusions, notably applicable to the Roman Empire.

%kobti 2012 la meme que kobti 2006 ?
%	pour 2006: etude de l'apparition d'inegalite

@@

To explore this coevolution between trade and cultural change we have developed an ABM where the different agents trade goods with variable prices.

\xrc{we can totally remove this paragraph and move the entire next subsection here:\\
These dynamics are modelled after processes (production, cultural transmission, innovation and trade) which will be explained in further details later. The flexibility of the models comes from the possibility to implement these processes differently depending on the hypothesis tested. However, since all experiments are performed with the same four processes, the effect of the various implementations of one process can be studied easily. Moreover, the operating of the model revolves around the value that agents associate to products. Depending on the question studied, the value can reflect the interest of the agent for the good or it can be the price at which the agent evaluate this good. Since this vector is common to all experiments performed with this model, it is extremely useful for the comparisons of results obtained with different implementation of the key processes.}

\xrc{This first paragraph should briefly explain the trade component?}

\subsection{Overview of the model}

@@

On top of these elements four processes are used: \emph{production}, \emph{cultural transmission}, \emph{innovation} and \emph{trade}. The \textit{production} process describes the creation of goods by the agent. Once a good is produced by an agent $i$ it is added its quantity vector ($Q^i$). The \emph{trade} process models the exchange of goods between the agents which results in a modification of the quantity vectors ($Q^i$). The amount of goods exchanged is computed by the agents concerned by the trade, within the \emph{trade} process, based on their value vectors ($V^i$). Within the \textit{cultural transmission} process an agent $i$ can copy the value vector ($V^j$) of an agent $j$, where $j \neq i$. The \textit{innovation} process also modifies the value vector $V^i$ of an agent, but it differs from the \emph{cultrual transmission} process in that the modification is done without reference to the other agents. Finally, .

This scheme achieves the two main goals listed in section~\ref{sec:intro}. The first goal is to be able to implement and compare various cultural transmission mechanisms as well as trade mechanisms within the same framework. We argue that this is possible in our framework by modifying only these four key processes i.e. production, trade, imitation and innovation. Moreover, thanks to the use of a unique $V^i$ vector, it is possible to compare the outcomes of different combinations of production, imitation and innovation mechanisms. The second goal is to produce a model composed of a low number of parameter. Our model is composed of only four processes and two vectors which is significantly less complex than previous models.

\xrc{I would remove the previous paragraph, because it is implied that we want few parameters and compare mechanisms}

The scheduling of the four processes (production, trade, cultural transmission and innovation) is described in algorithm~\ref{algo:complete} along with the vectors modified by each of these processes. Between line 2 and 7 all agents of the population are initialised with empty quantity vectors and random values. Between the lines 11 and 19 is presented the part of the pseudo-code describing the instructions that will be repeated for each iteration of the simulation. 
\jm{what about saying that we just count the iterations at the ``imitation, innovation step" ?}
\xrc{It should be justified (market is faster than cultural change?)}
One can note that the trading process is performed at every iteration while the imitation and innovation processes are executed only every $CulturalStep$. The idea behind this is to perform the imitation based on a score that reflects the performance of the agent and not only one lucky or unlucky trading round.

@@

\section{DISCUSSION}

In this paper, and thanks to the model developed, we have shown that it is the possible to implement trade mechanisms within a cultural evolutionary framework. In the end, it appears that the key aspects of trade are just different implementations of the key elements of cultural evolutionary framework, i.e. cultural transmission and innovation. Moreover, we have ensure that such an implementation is valid in the sense that it leads the prices of goods to converges to the values expected under economical theories. This aspect which has not been studied before, opens a new line of thinking on the weight of trade mechanisms within cultural change.

@@

\section{CONCLUSION}

@@

>>>>

\title{Modelling the co-evolution of trade and culture in the Roman market}

\maketitle

@@

% the spread of cultural traits is often modelled following neutral theory

Cultural change comprise the collection of processes that promotes or inhibits the spread of information by social interaction within a population~\shortcite[3]{boyd_origin_2005}. An increasing number of social scientists are using an evolutionary framework to model these dynamics, thus fostering the development of transdisciplinary efforts designed to understand cultural evolution~\shortcite{henrich_evolution_2003}.

Several studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependence properties. These analysis often compare evidence against neutral models~\shortcite{neiman_stylistic_1995}. They assume that traits do not affect the fitness of the individual that acquire it; within this context its transmission is unbiased, and its success will depend on its popularity amongst the rest of individuals. This neutral model generates a distinctive pattern of frequency distribution of traits, identified as a power law. It can be replicated with a simple random copying transmission mechanism~\shortcite{bentley_random_2004}: an individual will copy the traits of a randomly chosen individual with a given probability. This copy can potentially introduce some errors in the acquired trait, which account for innovation processes. The individual will in turn continue to spread these cultural traits which will be further adopted by other individuals. This basic model can be enriched by several additional processes both in the innovation~\shortcite{schillinger_copying_2014,sole_evolutionary_2013,ziman_technological_2003} and the transmissio~\shortcite{heyes_social_1994,henrich_evolution_2003}. Unbiased transmission works as a baseline for identifying frequency-dependent biases: if evidence has higher tendendy to copy the most common trait is is known as conformism, while the opposite is defined as anti-conformism.

This approach can be used as the starting point to analyse patterns of cultural change to explore frequency-dependance in archaeological contexts~\shortcite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010}. However, the fact that archaeology records fragmented material culture presents some challenges on the validity of the method~\shortcite{kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

This work explores the impact of a crucial element on the transmission of material culture: trade. Networks of good exchanges are being increasingly recognised as key elements that structured ancient societies~\shortcite{temin_market_2001,remesal_epnet_2014,brughmans_connecting_2010}. The scenarios where this process emerge suggest a complex bias in the selection of cultural traits, which at the same time are also identified as economic products~\shortcite{bentley_specialisation_2005,macmillan_agent-based_2008}. Trasmission is not neutral anymore, as different prices for each product will introduce a dynamic content bias. This affects the frequency of the product within the population, which in turn would modifies its price following a co-evolutionary dynamic.

This work explores these co-evolutionary dynamics using a theoretical Agent-Based Model (ABM), a type of simulation particularly useful for studying non-linear dynamics in heterogeneous environments within an evolutionary perspective~\cite{lake_trends_2014}. Next section defines the model, which is based on unbiased transmission mechanisms and simple exchange dynamics. Next, we define different experiments to explore the dynamics of the created model. Finally, the concluding remarks interpret the results and discuss further possibilities of the pesented model.

@@

@@

On top of these elements four processes are used: \emph{production}, \emph{cultural transmission}, \emph{innovation} and \emph{trade}. The \textit{production} process describes the creation of goods by the agent. Once a good is produced by an agent $i$ it is added its quantity vector ($Q^i$). The \emph{trade} process models the exchange of goods between the agents which results in a modification of the quantity vectors ($Q^i$). The amount of goods exchanged is computed by the agents concerned by the trade, within the \emph{trade} process, based on their value vectors ($V^i$). Within the \textit{cultural transmission} process an agent $i$ can copy the value vector ($V^j$) of an agent $j$, where $j \neq i$. The \textit{innovation} process also modifies the value vector $V^i$ of an agent, but it differs from the \emph{cultural transmission} process in that the modification is done without reference to the other agents. Finally, .

The scheduling of the four processes is described in algorithm~\ref{algo:complete} along with the vectors modified by each of these processes. Between line 2 and 7 all agents of the population are initialised with empty quantity vectors and random values. Between the lines 11 and 19 is presented the part of the pseudo-code describing the instructions that will be repeated for each iteration of the simulation. 
\jm{what about saying that we just count the iterations at the ``imitation, innovation step" ?}
\xrc{It should be justified (market is faster than cultural change?)}
One can note that the trading process is performed at every iteration while the imitation and innovation processes are executed only every $CulturalStep$. The idea behind this is to perform the imitation based on a score that reflects the performance of the agent and not only one lucky or unlucky trading round.

@@

\section{CONCLUDING REMARKS}

The results of this model show that the key aspects of trade are just different implementations of the key elements of cultural evolutionary framework, i.e. cultural transmission and innovation. Moreover, we have ensure that such an implementation is valid in the sense that it leads the prices of goods to converges to the values expected under economical theories. This aspect which has not been studied before, opens a new line of thinking on the weight of trade mechanisms within cultural change.

@@

@@

\xrc{I think that this section can be revised once we get the results (e.g. less future work, more interpretation of results)}
